### 2018-02-10 results.tex
prepared verification test figures

Some aspects of Fig.~\ref{fig:clusters} are worth discussing. Although in most plots all four clusters are clearly distinguishable, there are others in which that is not the case. In the C5-C9 and C7-C9 combinations in the 3-feature analysis, for instance, the excellent cluster has a limited presence. This is due to the influence of the lower C9 values (see Fig.~\ref{fig:data-box-plot}) on the correlations with C5 and C7. This is not necessarily a bad thing, as it helps to understand the different relationships between the various GOF metrics. In the best of cases, the combinations C1-C2 and C5-C8 in the 2- and 4-feature analysis, for instance, exhibit an almost perfect proportionality between the metrics. These are called redundant relationships, where knowledge about a feature provides a direct view into another feature's results. On the other hand, the combination C1-C7 in the 2-feature analysis presents an example of an irrelevant relationship, where knowledge about a feature provides no insight about the other. Subspaces with redundant relationships are useful for reducing the number of features. Irrelevant relationships, however, do not contribute much to the whole process \citep{Dy_2004_MLR}.

\begin{figure*}[ht!]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-05}
	\caption{A sample of the results from the multi-dimensional clustering analysis showing the clusters from a two-dimensional perspective into the relationships between different GOF metrics. In each case, the labels near the upper-left corner indicate the features considered in the subspace analysis, and the features associated to the horizontal ($x$) and vertical ($y$) axes in the plot. The top, middle, and bottom rows correspond to 2-, 3-, and 4-feature subspace analyses. In each case, the poor, fair, good, and excellent clusters are indicated with circle, triangle, square, and cross symbols, respectively. The color version of this figure is available only in the electronic edition.}
	\label{fig:clusters}
\end{figure*}

Figure \ref{fig:boxed-clusters} shows the statistical distribution (box-plots) of the samples once the dataset is partitioned into the four validation classes. This is similar to Fig.~\ref{fig:data-box-plot}, but after the clustering process is completed. Separately, we prepared similar plots to look at the influence of the velocity models and components of motions on the results of the clustering process and, as observed before, they had no significant differences with the aggregate of all the samples in the dataset.

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-06}
	\caption{Statistical distribution of the dataset as partitioned into the four validation categories (poor, fair, good, excellent) after the clustering analysis. The distributions are shown in the form of box-plots for each metric (C1 through C11), and discretized in terms of the component of motion (NS, EW, UD) for visualization purposes. In each case, the median is indicated by a notch in the box of the central quartiles, and the lines represent the interquartile range, with outliers shown as scattered dots. The color version of this figure is available only in the electronic edition.}
	\label{fig:boxed-clusters}
\end{figure*}

In total, the clustering process results in 816, 1253, 879 and 76 data samples for the poor, fair, good and excellent classes, respectively. These groups are shown in Fig.~\ref{fig:count-classes}. As it can be seen in this figure, the number of samples in the excellent class is significantly less than those in the poor, fair, and good classes. Therefore, before moving on with the decision tree analysis, it is necessary to resample the subset of the excellent class. We used the oversampling approach described in the previous section. We replicated data from the excellent class randomly by a $\times$10 factor until elevating the number of samples in this class to 760 as indicated in Fig.~\ref{fig:count-classes} with the dashed line. By common standards, an oversample rate of $\times$10 is considered acceptable \citep{Weiss_2003_JAIR}. (Arguably, we could have as well undersampled the fair class, but we deemed that unnecessary.) Once this process was completed we went on with the decision tree analysis.

\begin{figure}[t]
	\centering
	\includegraphics[width=\columnwidth]{figures/pdf/figure-07}
	\caption{Number of data samples in each class (poor, far, good, excellent) after conducting a multi-dimensional constrained \kmeans{} clustering process using subspace analysis for 2, 3, and 4 features (GOF metrics). Here, for illustrative purposes, the filling of the bars distinguishes the component of motion of the compared signals as classified. The dashed-line bar indicates the number of samples in the excellent class after oversampling. The color version of this figure is available only in the electronic edition.}
	\label{fig:count-classes}
\end{figure}

@@

For each one of these unique trees, we computed the effectiveness factor $F_1$ from equation (\ref{eq:f}), and extracted the total number of nodes in the trees and their depth. Figure \ref{fig:nodes-depths} shows the results of $F_1$ for all the trees and its distribution in terms of the number of attributes used in the trees as a function of the number of nodes and the depth of each tree. Recall that we are interested in finding a sequence of decisions (represented by disjunctive decision nodes in a tree) that can lead to good GOF predictions (i.e, high values of $F_1$) using a reduced number of attributes (GOF metrics). In general, all the trees obtained with the C5.0 algorithm are good in terms of the effectiveness factor ($F_1$ values close to 1). Then our choice comes down to using a reduced number of metrics. Having several trees with 2, 3, and 4 metrics (as opposed to 11), the following factor in the decision is choosing trees with algorithms using a small number of steps to reach the prediction. This is given by a combination between the depth of the tree and the number of nodes in the tree (i.e., trees with low complexity).

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-08}
	\caption{Trees effectiveness results in terms of the factor ($F_1$) from equation \ref{eq:f} as indicated by colored dots distributed by the number of attributes (GOF metrics) as a function of \textbf{(a)} the number of nodes, and \textbf{(b)} the depth of the trees. The rings around the left-most bottom dots indicate selected trees shown in Fig.~\ref{fig:trees}. The color version of this figure is available only in the electronic edition.}
	\label{fig:nodes-depth}
\end{figure*}

Based on these criteria, we selected three candidate trees: T1, T2, and T3. These trees are shown in Fig.~\ref{fig:trees}. T1 is the simplest of the three, and T2 and T3 share some of the logarithmic structure on the branches to the right-hand side. T3 is the more complex of them. More complex trees tend to be deeper, use more nodes, and thus employ more attributes. In total T1, T2, and T3 use 2, 3 and 4 attributes respectively. All coincide in the use of the total energy (C4) and response spectra (C8) as key metrics. T2 adds peak acceleration (C5), and T3 adds peak velocity (C5) to the previous metrics.

>>>>

Some aspects of Fig.~\ref{fig:clusters} are worth discussing. Although in most plots all four clusters are clearly distinguishable, there are others in which that is not the case. In the C5-C9 and C7-C9 combinations in the 3-feature analysis, for instance, the excellent cluster has a limited presence. This is due to the influence of the lower C9 values (see Fig.~\ref{fig:data-box-plot}) on the correlations with C5 and C7. This is not necessarily a bad thing, as it helps to understand the different relationships between the various GOF metrics. In the best of cases, the combinations C1-C2 and C5-C8 in the 2- and 4-feature analysis, for instance, exhibit an almost perfect proportionality between the metrics. That means either one of the metrics in those combinations is redundant. Relationships with redundant features are those where knowledge about one of the features provides a direct view into the other. On the other hand, the combination C1-C7 in the 2-feature analysis presents an example of an irrelevant feature. We say C1 is irrelevant because it provides no insight about the outcome of C7. Identifying subspaces with redundant and irrelevant features is important because, on the one hand, the former help reduce the number of necessary features, while on the other hand, the latter can essentially be discarded because of their weak contribution to the decision making process \citep{Dy_2004_MLR}.

\begin{figure*}[ht!]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-05}
	\caption{A sample of the results from the multi-dimensional clustering analysis showing the clusters from a two-dimensional perspective into the relationships between different GOF metrics. In each case, the labels near the upper-left corner indicate the features considered in the subspace analysis, and the features associated to the horizontal ($x$) and vertical ($y$) axes in the plot. The top, middle, and bottom rows correspond to 2-, 3-, and 4-feature subspace analyses. In each case, the poor, fair, good, and excellent clusters are indicated with circle, triangle, square, and cross symbols, respectively. The empty circles indicate the location of the artificial cannot-link stations we introduced as background knowledge to the clustering process. The color version of this figure is available only in the electronic edition.}
	\label{fig:clusters}
\end{figure*}

Figure \ref{fig:boxed-clusters} shows the statistical distribution (box-plots) of the samples once the dataset is partitioned into the four GOF validation classes. This is similar to Fig.~\ref{fig:data-box-plot}, but after the clustering process is completed. Separately, we prepared similar plots to look at the influence of the velocity models and components of motions on the results of the clustering process and, as observed before, they had no significant differences with the aggregate of all the samples in the dataset.

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-06}
	\caption{Statistical distribution of the dataset as partitioned into the four validation categories (poor, fair, good, excellent) after the clustering analysis. The distributions are shown in the form of box-plots for each metric (C1 through C11). In each case, the median is indicated by a notch in the box of the central quartiles, and the lines represent the interquartile range, with outliers shown as scattered dots. The color version of this figure is available only in the electronic edition.}
	\label{fig:boxed-clusters}
\end{figure*}

In total, the clustering process results in 816, 1253, 879 and 76 data samples for the poor, fair, good and excellent classes, respectively. These groups are shown in Fig.~\ref{fig:count-classes}. As it can be seen in this figure, the number of samples in the excellent class is significantly less than those in the poor, fair, and good classes. Therefore, before moving on with the decision tree analysis, it is necessary to resample the subset of the excellent class. We used the oversampling approach described in the previous section. We replicated data from the excellent class randomly by a $\times$10 factor until elevating the number of samples in this class to 760 as indicated in Fig.~\ref{fig:count-classes} with the dashed line. By common standards, an oversample rate of $\times$10 is considered acceptable \citep{Weiss_2003_JAIR}. (Arguably, we could have as well undersampled the fair class, but we deemed that unnecessary. Instead, we used a strong pruning process to prevent overfitting.) Once this process was completed we went on with the decision tree analysis.

\begin{figure}[t]
	\centering
	\includegraphics[width=\columnwidth]{figures/pdf/figure-07}
	\caption{Number of data samples in each class (poor, fair, good, excellent) after conducting a multi-dimensional constrained \kmeans{} clustering process using subspace analysis for 2, 3, and 4 features (GOF metrics). The dashed-line bar indicates the number of samples in the excellent class after oversampling. The color version of this figure is available only in the electronic edition.}
	\label{fig:count-classes}
\end{figure}

@@

For each one of these unique trees, we computed the effectiveness factor $F_1$ from equation (\ref{eq:f}), and extracted the total number of nodes in the trees and their depth. Figure \ref{fig:nodes-depth} shows the results of $F_1$ for all the trees and its distribution in terms of the number of attributes used in the trees as a function of the number of nodes and the depth of each tree. Recall that we are interested in finding a sequence of decisions (represented by disjunctive decision nodes in a tree) that can lead to good GOF predictions (i.e, high values of $F_1$) using a reduced number of attributes (GOF metrics). In general, all the trees obtained with the C5.0 algorithm are good in terms of the effectiveness factor ($F_1$ values close to 1). Then our choice comes down to using a reduced number of metrics. Having several trees with 2, 3, and 4 metrics (as opposed to 11), the following factor in the decision is choosing trees with algorithms using a small number of steps to reach the prediction. This is given by a combination between the depth of the tree and the number of nodes in the tree (i.e., trees with low complexity).

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-08}
	\caption{Accuracy of tree predictions in terms of the factor ($F_1$) from equation \ref{eq:f} as indicated by colored dots distributed with respect to the number of attributes (GOF metrics) as a function of: \textbf{(a)} the number of nodes, and \textbf{(b)} the depth of the trees. The rings around the left-most bottom dots indicate selected trees shown in Fig.~\ref{fig:trees}. The color version of this figure is available only in the electronic edition.}
	\label{fig:nodes-depth}
\end{figure*}

Based on these criteria, we selected three candidate trees: T1, T2, and T3. These trees are shown in Fig.~\ref{fig:trees}. T1 is the simplest of the three, and T2 and T3 share some of their topology on the right-hand side. T3 is the most complex of them. More complex trees tend to be deeper, have more nodes, and employ more attributes. This leads to higher $F_1$ values, but they may not necessarily be more practical. In total T1, T2, and T3 use 2, 3 and 4 attributes respectively. All coincide in the use of the total energy (C4) and response spectra (C8) as key metrics. T2 adds peak acceleration (C5), and T3 adds peak velocity (C5) to the previous metrics.
